Global phase diagram and possible quantum spin liquid in the triangular $J$-$K$-$\Gamma$ model

### 2020-11-11 paper/paper.tex
Paper 2020-08-21 15:55

Geometric frustration, which arises when the lattice geometry gives rise to constraints that not every exchange bond can be simultaneously minimized in energy, plays an important role in various kinds of magnetic systems. The nearest-neighbor (NN) antiferromagetic (AFM) Heisenberg model on the triangular lattice is a typical example, once two of the spins on an elementary triangle are antiparallel to satisfy their antiferromagnetic interaction, the third one can no longer point in a direction opposite to both other spins. In particular, the spin-1/2 case has attracted numerous interests and was extensively studied after the seminal prediction by Anderson of a ``quantum spin liquid" (QSL) where a strong quantum fluctuation prevents any long-range order down to the zero temperature~\cite{Anderson1973}, although it has been shown by later studies that the ground state (GS) have a classical magnetic order with each spin on a triangle pointing to 120$^\circ$ with respect to each other~\cite{PhysRevLett.99.127004,PhysRevLett.82.3899,PhysRevB.50.10048,PhysRevLett.60.2531}. When the second NN interactions are included which introduce further frustration, the system has a much richer phase diagram including the 120$^\circ$ N\'{e}el state, the stripe state and a QSL state~\cite{PhysRevB.91.014426,PhysRevB.92.041105,PhysRevB.92.140403,PhysRevB.96.165141,PhysRevB.93.144411,JPSJ.83.093707,PhysRevB.94.121111,PhysRevB.96.075116,PhysRevLett.120.207203,PhysRevLett.123.207203}. All these studies have revealed that geometric frustated systems show quite different behavior from that of the non-frustated system.

On the other hand, exchange frustation in systems with strongly anisotropic magnetic interactions has been shown to be another promising approach to explore exotic quantum spin states. Like geometric frustation, the effect of exchange frustration is to prevent the formation of long range magnetic order and give raise to a residual ground-state entropy. The spin-$1/2$ Kitaev model~\cite{Kitaev2006} on honeycomb lattice, which has both gapped and gapless QSL states supporting fractionalized excitations, is a celebrated example of a model with exchange frustration. In this model, the spins subject to the Kitaev interactions consisting of nearest neighbor Ising-type interactions, with the quantization axis depending on the spatial orientation of an exchange bond. Because of its theoretical importance and potential application in quantum computing, great efforts have been made to search for a solid-state realization of the Kitaev model. G. Khaliullin \emph{et al.,}~\cite{Khaliullin2005,PhysRevLett.102.017205} proposed that this highly anisotropic Kitaev interaction can be realized in 4d/5d systems with a low spin state of $d^5$ configuration, such as iridates A$_2$IrO$_3$ (A = Na, Li). In these systems, the bond-directional interactions originate from the joint effects of strong spin-orbital coupling (SOC), electron interactions, $d^5$ configuration and 90$^\circ$ bond geometry formed by edge sharing octahedra. However, in real materials, other types of interaction besides the Kitaev coupling are present, and these interactions may induce other interesting ordered and disordered phases. The simplest extension of the pure Kitaev model is the Heisenberg-Kitaev (HK) model~\cite{PhysRevLett.105.027204}, in which the NN Heisenberg interaction is also taken into account. This model has been extensively studied by various numerical methods~\cite{PhysRevLett.110.097204,PhysRevB.83.245104,PhysRevB.84.100406,PhysRevB.90.195102,PhysRevLett.119.157203}, which reveal the presence of four magnetically ordered phases with collinear spin patterns of ferromagnetic (FM), AFM, stripe, and zigzag types, besides extended spin-liquid phases near the Kitaev limits. Considering the most idealized crystal structure, another interaction beyond the HK model also must be included, i.e. bond dependent symmetric off-diagonal exchange, which is called the $\Gamma$ interaction~\cite{PhysRevLett.112.077204,PhysRevB.93.214431,PhysRevB.96.115103}. Thus, the generic NN exchange Hamiltonian for the undistorted hexagonal compounds is the $J$-$K$-$\Gamma$ model, where the Heisenberg ($J$), Kitaev ($K$) and $\Gamma$ interactions are all included. Finite $\Gamma$ further enriches the phase diagram by adding non-collinear and incommensurate spiral phases~\cite{PhysRevLett.112.077204}. Moreover, in real materials, such as A$_2$IrO$_3$ and $\alpha$-RuCl$_{3}$, the dominant interactions are the $\Gamma$ and FM Kitaev terms which originate from both direct $d$-$d$ and anion mediated $d$-$p$ electron transfer, while the Heisenberg term has the smallest strength since it predominantly originates from the weak direct $d$-$d$ electron transfer~\cite{PhysRevLett.112.077204,PhysRevB.93.214431,PhysRevB.96.115103}.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/ModelDefinition.pdf}
    \caption{\label{fig:ModelDefinition}(Color online) (a) Top view of the triangular lattice of the edge-sharing octahedron. (b) The orientation of the cubic $x$, $y$, $z$ axes with respect to the octahedron. The spin operators $S^x$, $S^y$ and $S^z$ are defined with respect to this reference frame. (c) Three types of the NN bonds on the triangular lattice, namely $\gamma=x, y, z$ colored red, green and blue, respectively. The different colors of the lattice sites label the four sublattices realizing the four sublattice transformation ( see main text for details).}
\end{figure}
In fact, magnetic ions located at the center of edge-sharing octahedra can not only form the honeycomb lattice but also the triangular lattice (see Fig.~\ref{fig:ModelDefinition}(a)), so the Kitaev and $\Gamma$ terms can naturally be generalized to the triangular lattice~\cite{PhysRevB.89.014414,PhysRevB.93.104417}. On the experimental side, studies on several classes of  compounds containing localized $4d$, $5d$ or $4f$ electrons have recently shown that the quantum spin model on the triangular lattice can be formed by the localized moments~\cite{PhysRevB.86.140405,PhysRevB.94.174410,srep16419,PhysRevLett.115.167203,PhysRevLett.120.087201}. In particular, due to their possible Kramers doublets and spin-orbital coupling, these moments can be treated as $S_{eff}=1/2$ at low temperatures, and the spin-orbital entanglement can induce direction-dependent exchanges, such as the $K$ and $\Gamma$ terms. A typical example is YbMgGaO$_{4}$~\cite{srep16419,PhysRevLett.115.167203}, in which the Yb$^{3+}$ ions form a triangular layer and are surrounded by O$^{2-}$ which construct edge sharing octahedra. More recently, an alternative family of compounds AReCh$_2$ (A=alkali, Re=rare-earth, Ch=O, S, Se) with perfect triangular lattices of rare-earth ions have been synthesized and explored. The magnetic susceptibilty and heat capacity data suggest no long-range magnetic order or spin freezing down to the lowest measurement temperature, which implies their candidacy for QSL state~\cite{acsmaterialslett.9b00464,PhysRevMaterials.3.114413,PhysRevB.100.241116,Liu_2018,PhysRevB.99.180401,PhysRevB.100.224417,PhysRevB.100.144432}. These triangluar magnets provide suitable platforms to study the interplay of geometric frustation and exchange frustration induced by spin-orbital couplings.

@@

In this paper, we study the global phase diagram and the phase transitions of the triangular lattice $J$-$K$-$\Gamma$ model in detail using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin directions for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-B and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $\Gamma=\pm1$ ($J=K=0$) in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the numerical simulations by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state. For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, other long-range ordered states which are energetically degenerate with the FM state also appear (see Appendix~\ref{apx:DegeneratedStates} for details). However, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states according to our ED calculations. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) (a)-(h) Representative SSFs from the ED calculations for different interaction parameters. The green dashed lines marks the first BZ corresponding to the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}

>>>>

Geometric frustration, which arises when the lattice geometry gives rise to constraints that not every exchange bond can be simultaneously minimized in energy, plays an important role in various kinds of magnetic systems. The nearest-neighbor (NN) antiferromagetic (AFM) Heisenberg model on the triangular lattice is a typical example, once two of the spins on an elementary triangle are antiparallel to satisfy their antiferromagnetic interaction, the third one can no longer point in a direction opposite to both other spins. In particular, the spin-1/2 case has attracted numerous interests and was extensively studied after the seminal prediction by Anderson of a ``quantum spin liquid" (QSL) where a strong quantum fluctuation prevents any long-range order down to the zero temperature~\cite{Anderson1973}, although it has been shown by later studies that the ground state (GS) have a classical magnetic order with each spin on a triangle pointing to 120$^\circ$ with respect to each other~\cite{PhysRevLett.99.127004,PhysRevLett.82.3899,PhysRevB.50.10048,PhysRevLett.60.2531}. When the interactions beyond the NN Heisenberg type are included which introduce further frustration, the system has a much richer phase diagram including the 120$^\circ$ N\'{e}el state, the stripe state and a QSL state~\cite{PhysRevB.91.014426,PhysRevB.92.041105,PhysRevB.92.140403,PhysRevB.96.165141,PhysRevB.93.144411,JPSJ.83.093707,PhysRevB.94.121111,PhysRevB.96.075116,PhysRevLett.120.207203,PhysRevB.95.165110,PhysRevLett.123.207203,PhysRevX.9.031026}. All these studies have revealed that geometric frustated systems show quite different behavior from that of the non-frustated system.

On the other hand, exchange frustation in systems with strongly anisotropic magnetic interactions has been shown to be another promising approach to explore exotic quantum spin states. Like geometric frustation, the effect of exchange frustration is to prevent the formation of long range magnetic order and give raise to a residual ground-state entropy. The spin-$1/2$ Kitaev model~\cite{Kitaev2006} on honeycomb lattice, which has both gapped and gapless QSL states supporting fractionalized excitations, is a celebrated example of a model with exchange frustration. In this model, the spins subject to the Kitaev interactions consisting of nearest neighbor Ising-type interactions, with the quantization axis depending on the spatial orientation of an exchange bond. Because of its theoretical importance and potential application in quantum computing, great efforts have been made to search for a solid-state realization of the Kitaev model. G. Khaliullin \emph{et al.,}~\cite{Khaliullin2005,PhysRevLett.102.017205} proposed that this highly anisotropic Kitaev interaction can be realized in 4d/5d systems with a low spin state of $d^5$ configuration, such as iridates A$_2$IrO$_3$ (A = Na, Li). In these systems, the bond-directional interactions originate from the joint effects of strong spin-orbital coupling (SOC), electron interactions, $d^5$ configuration and 90$^\circ$ bond geometry formed by edge sharing octahedra. However, in real materials, other types of interaction besides the Kitaev coupling are present, and these interactions may induce other interesting ordered and disordered phases. The simplest extension of the pure Kitaev model is the Heisenberg-Kitaev (HK) model~\cite{PhysRevLett.105.027204,10.1146/annurev-conmatphys-033117-053934}, in which the NN Heisenberg interaction is also taken into account. This model has been extensively studied by various numerical methods~\cite{PhysRevLett.110.097204,PhysRevB.83.245104,PhysRevB.84.100406,PhysRevB.90.195102,PhysRevLett.119.157203}, which reveal the presence of four magnetically ordered phases with collinear spin patterns of ferromagnetic (FM), AFM, stripe, and zigzag types, besides extended spin-liquid phases near the Kitaev limits. Considering the most idealized crystal structure, another interaction beyond the HK model also must be included, i.e. bond dependent symmetric off-diagonal exchange, which is called the $\Gamma$ interaction~\cite{PhysRevLett.112.077204,PhysRevB.93.214431,PhysRevB.96.115103}. Thus, the generic NN exchange Hamiltonian for the undistorted hexagonal compounds is the $J$-$K$-$\Gamma$ model, where the Heisenberg ($J$), Kitaev ($K$) and $\Gamma$ interactions are all included. Finite $\Gamma$ further enriches the phase diagram by adding non-collinear and incommensurate spiral phases~\cite{PhysRevLett.112.077204,Winter_2017,10.1038/s42254-019-0038-2}. Moreover, in real materials, such as A$_2$IrO$_3$ and $\alpha$-RuCl$_{3}$, the dominant interactions are the $\Gamma$ and FM Kitaev terms which originate from both direct $d$-$d$ and anion mediated $d$-$p$ electron transfer, while the Heisenberg term has the smallest strength since it predominantly originates from the weak direct $d$-$d$ electron transfer~\cite{PhysRevLett.112.077204,PhysRevB.93.214431,PhysRevB.96.115103}.

\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/ModelDefinition.pdf}
    \caption{\label{fig:ModelDefinition}(Color online) (a) Top view of the triangular lattice of the edge-sharing octahedron. (b) The orientation of the cubic $x$, $y$, $z$ axes with respect to the octahedron. The spin operators $S^x$, $S^y$ and $S^z$ are defined with respect to this reference frame. (c) Three types of the NN bonds on the triangular lattice, namely $\gamma=x, y, z$ colored red, green and blue, respectively. The different colors of the lattice sites label the four sublattices realizing the four sublattice transformation ( see main text for details).}
\end{figure}
In fact, magnetic ions located at the center of edge-sharing octahedra can not only form the honeycomb lattice but also the triangular lattice (see Fig.~\ref{fig:ModelDefinition}(a)), so the Kitaev and $\Gamma$ terms can naturally be generalized to the triangular lattice~\cite{PhysRevB.89.014414,PhysRevB.93.104417}. On the experimental side, studies on several classes of  compounds containing localized $4d$, $5d$ or $4f$ electrons have recently shown that the quantum spin model on the triangular lattice can be formed by the localized moments~\cite{PhysRevB.86.140405,PhysRevB.94.174410,srep16419,PhysRevLett.115.167203,PhysRevLett.120.087201,CEVALLOS2018154,10.1038/s41535-019-0151-6}. In particular, due to their possible Kramers doublets and spin-orbital coupling, these moments can be treated as $S_{eff}=1/2$ at low temperatures, and the spin-orbital entanglement can induce direction-dependent exchanges, such as the $K$ and $\Gamma$ terms. A typical example is YbMgGaO$_{4}$~\cite{srep16419,PhysRevLett.115.167203}, in which the Yb$^{3+}$ ions form a triangular layer and are surrounded by O$^{2-}$ which construct edge sharing octahedra, but due to the inherent disorder effect there are still controversies about the GS of this material~\cite{srep16419,PhysRevLett.115.167203,PhysRevLett.120.087201,PhysRevB.94.035107,Nature20614,nphys3971,PhysRevLett.117.267202,PhysRevX.8.031028,SciPostPhys.4.1.003,PhysRevB.97.184413}. More recently, an alternative family of compounds AReCh$_2$ (A=alkali, Re=rare-earth, Ch=O, S, Se) with perfect triangular lattices of rare-earth ions have been synthesized and explored. The magnetic susceptibilty and heat capacity data suggest no long-range magnetic order or spin freezing down to the lowest measurement temperature, which implies their candidacy for QSL state~\cite{acsmaterialslett.9b00464,PhysRevMaterials.3.114413,PhysRevB.100.241116,Liu_2018,PhysRevB.99.180401,PhysRevB.100.224417,PhysRevB.100.144432}. These triangluar magnets provide suitable platforms to study the interplay of geometric frustation and exchange frustration induced by spin-orbital couplings.

@@

In this paper, we comprehensively study the global phase diagram and the phase transitions of the triangular lattice $J$-$K$-$\Gamma$ model using a combination of ED and classical analyses. We find that the ED results are basically consistent with the classical analyses such as the classical Monte Carlo simulations. In the HK limit ($\Gamma=0$), there are five classical phases: one FM, two stripe, one $120^{\circ}$ N\'{e}el and its dual phases. For the pure AFM Kitaev model, although the system has highly degenerate classical GSs composed of the stripe and nematic states, the order-by-disorder mechanism caused by quantum fluctuations makes the system select the stripe state to be the GS. When the $\Gamma$ term is included, we find five new phases: two FM, one stripe, one modulated stripe and one possible QSL phases. With the aid of classical analysis, we determine the spin configuration for each magnetically ordered phase and explain that why there are phase transitions between the phases with the same type of classical orders. On the other hand, we find that the order-by-disorder mechanism makes the GSs of the pure $\Gamma$ models have FM orders, although the classical analyses suggest that the GSs are highly degenerate. For the possible QSL phase, based on the investigation of the spin excitation spectrum, we find that the GS is probably a gapped $Z_{2}$ QSL.

@@

For the other two phases with large areas and mainly located in the $J<0$ region of the phase diagram, we show the SSFs of two representative points with $\alpha=0.3\pi, \beta=1.3\pi$ and $\alpha=0.7\pi, \beta=1.65\pi$ in Fig.~\ref{fig:StructureFactors}(f) and (g), respectively. Since the SSFs are peaked at $\tilde{\Gamma}$ point of the first BZ, so both phases have ferromagnetic orders and they are denoted with FM-B and FM-C in Fig.~\ref{fig:QuantumPhaseDiagram}. We note that there are two special points with $\Gamma=\pm1$ ($J=K=0$) in the two FM phases, which are located at the phase boundary in the classical phase diagram~\cite{PhysRevB.92.165108}. Thus, it is worthy to discuss the properties of the pure $\Gamma$ models in details, and we perform the classical analyses by using the ``basin-hopping" global optimization and classical Monte Carlo methods. For $\Gamma=+1$, apart from the FM GS with the spin orientation lying in the lattice plane, we also found disordered states having the same energy as the FM state (see Appendix~\ref{apx:DegeneratedStates} for details). For $\Gamma=-1$, in addition to the FM order perpendicular to the lattice plane, there are also several noncollinear magnetic orders energetically degenerate with the FM state (see Appendix~\ref{apx:DegeneratedStates} for details). However, according to our ED calculations, for the quantum model, quantum flucations select the FM ordered states as the GS out of the degenerate manifolds of classical states. This selection of states among the degenerate classical ground states is the so the called order-by-disorder mechanism, which has previously been applied in a number of insulating magnets, especially in cases where frustration leads to a degenerate manifold of classical ground state configurations that is broken by quantum fluctuations~\cite{JPSJ.54.4494,PhysRevLett.62.2056,PhysRevB.46.11137,PhysRevLett.88.067203,PhysRevB.81.214419,PhysRevLett.109.077204,PhysRevX.5.041035}.
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{fig/StructureFactors.pdf}
    \caption{\label{fig:StructureFactors}(Color online) (a)-(h) Representative SSFs from the ED calculations for different interaction parameters. The green dashed lines marks the first BZ corresponding to the triangular lattice. (i) Characteristic wave vectors for the FM and stipe phases. See Appendix~\ref{apx:SSF} for the SSFs of the 120$^\circ$ N\'{e}el phase, dual N\'{e}el phase and pure $\Gamma$ models.}
\end{figure}
